Nonparametric Clustering Based on Energy Distance

### 2017-05-04 qcqp_clustering/draft/energy_clustering.tex
draft

\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}

%\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
\usepackage{bbm}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

@@

\author{Guilherme Fran\c ca}
\email{guifranca@gmail.com}
\affiliation{Johns Hopkins University, Center for Imaging Science}

\author{Joshua Vogelstein}
\email{jovo@jhu.edu}
\affiliation{Johns Hopkins University, Center for Imaging Science}

@@

In this section we briefly review the main concepts from energy
statistics and reproducing kernel Hilbert spaces (RKHS) used in this paper. 
For more details we refer the reader
to \cite{Szkely2013} (and references therein) and also \cite{Sejdinovic2013}.

@@

There is an equivalence 
between energy distance, commonly used in statistics,
and distances between embeddings of distributions in reproducible
kernel Hilbert spaces (RKHS),
commonly used in machine learning. This equivalence was established
in \cite{Sejdinovic2013}. We recall the definition of
RKHS. Let $\mathcal{H}$ be a Hilbert space over real-valued functions
over $\mathbb{R}^D$. A function $k : \mathbb{R}^D \times \mathbb{R}^D \to 
\mathbb{R}$ is a reproducing kernel of $\mathcal{H}$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $K_x \equiv k(\cdot, x) \in \mathcal{H}$ for any $x \in \mathbb{R}^D$.
\item $\langle K_x, f \rangle_{\mathcal{H}} = f(x)$ for
any $x\in\mathbb{R}^D$ and $f\in \mathcal{H}$.
\end{enumerate}
In other words, for any $x \in \mathbb{R}^D$ there is a unique function
$K_x \in \mathcal{H}$ that reproduces $f(x)$ through the inner product
of $\mathcal{H}$.
If such a kernel function $k$ exists, then $\mathcal{H}$ is called a RKHS.
From this we have $\langle K_x, K_y \rangle = K_y(x) = k(x,y)$. This implies
that $k(x,y)$ is symmetric and positive definite, $\sum_{i,j=1}^n c_i c_j
k(x_i,x_j) \ge 0$ for $c_i,c_j \in \mathbb{R}$.

@@

\section{Clustering using Energy Distance}

>>>>

%\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
\documentclass{article}

%\usepackage{nips_2017}

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2017}

\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[inline]{enumitem}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figs/}}

%\usepackage{cite}
\usepackage{bbm}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\author{
Guilherme Fran\c ca \\
Johns Hopkins University\\
\texttt{guifranca@gmail.com} \\
%% examples of more authors
\And
Joshua Vogelstein \\
Johns Hopkins University\\
\texttt{jovo@jhu.edu}
}

%\author{Guilherme Fran\c ca}
%\email{guifranca@gmail.com}
%\affiliation{Johns Hopkins University, Center for Imaging Science}

%\author{Joshua Vogelstein}
%\email{jovo@jhu.edu}
%\affiliation{Johns Hopkins University, Center for Imaging Science}

@@

@@

In this section we briefly review the main concepts from energy
statistics and reproducing kernel Hilbert spaces (RKHS) used through paper. 
For more details we refer the reader
to \cite{Szkely2013} 
(and references therein) and also \cite{Sejdinovic2013}.

@@

There is an equivalence 
between energy distance, commonly used in statistics,
and distances between embeddings of distributions in reproducing
kernel Hilbert spaces (RKHS),
commonly used in machine learning. This equivalence was established
in \cite{Sejdinovic2013}. We recall the definition of
RKHS. Let $\mathcal{H}$ be a Hilbert space over real-valued functions
over $\mathbb{R}^D$. A function $k : \mathbb{R}^D \times \mathbb{R}^D \to 
\mathbb{R}$ is a reproducing kernel of $\mathcal{H}$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $K_x \equiv k(\cdot, x) \in \mathcal{H}$ for any $x \in \mathbb{R}^D$.
\item $\langle K_x, f \rangle_{\mathcal{H}} = f(x)$ for
any $x\in\mathbb{R}^D$ and $f\in \mathcal{H}$.
\end{enumerate}
In other words, for any $x \in \mathbb{R}^D$ there is a unique function
$K_x \in \mathcal{H}$ that reproduces $f(x)$ through the inner product
of $\mathcal{H}$.
If such a kernel function $k$ exists, then $\mathcal{H}$ is called a RKHS.
From this we have $\langle K_x, K_y \rangle = K_y(x) = k(x,y)$. This implies
that $k(x,y)$ is symmetric and positive definite, $\sum_{i,j=1}^n c_i c_j
k(x_i,x_j) \ge 0$ for $c_i,c_j \in \mathbb{R}$.

@@

\section{Energy Distance based Clustering}

Now we formulate an optimization problem for clustering based on 
energy statistics
and RKHS.
Assume we have data $\mathcal{X} = \{ x_1,x_2,\dotsc, x_N \}$,
and a partition $\mathcal{X} = \cup_{k=1}^K \mathcal{A}_k$, where
$\mathcal{A}_k \cap \mathcal{A}_{k'} = \emptyset$.
Each expectation in \eqref{eq:energy2} can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g_\alpha(\mathcal{A}_k, \mathcal{A}_{k'}) \equiv 
\dfrac{1}{N_k N_{k'}}\sum_{x \in \mathcal{A}_k} 
\sum_{y \in \mathcal{A}_{k'}} \| x - y\|^\alpha
\end{equation}
where $N_k = |\mathcal{A}_k|$ is the number of elements in partition
$\mathcal{A}_k$. In energy statistics 
\cite{Szkely2013} we have the within energy dispersion
\begin{equation}
\mathcal{W}_\alpha = 
\sum_{k=1}^{K} \dfrac{N_k}{2} g_\alpha(\mathcal{A}_k, \mathcal{A}_k),
\end{equation}
and also the between-sample energy statistic
\begin{equation}
\mathcal{S}_\alpha = 
\sum_{1 \le  k < k' \le K } \dfrac{N_k N_{k'}}{2 N} \left[
2 g_{\alpha}(\mathcal{A}_k, \mathcal{A}_{k'}) - 
g_{\alpha}(\mathcal{A}_k, \mathcal{A}_{k}) - 
g_{\alpha}(\mathcal{A}_{k'}, \mathcal{A}_{k'})
\right].
\end{equation}
This is a test statistic for equality of distributions, which is small
if all datapoints comes from the same distribution, and diverges otherwise,
in the limit $N \to \infty$. Therefore, our criteria for 
clustering data is to 
maximize $\mathcal{S}_\alpha$.
It can be shown that the total dispersion of the data obeys \cite{Szkely2013}
\begin{equation}
\mathcal{T}_\alpha(\mathcal{X}) 
= \mathcal{W}_\alpha + \mathcal{S}_\alpha = \dfrac{N}{2}
g_{\alpha}(\mathcal{X}, \mathcal{X}). 
\end{equation}
Note that $\mathcal{T}$ only depends on the pooled data, so it 
does not depend on how we partition $\mathcal{X}$. Therefore, maximizing
$\mathcal{S}_\alpha$ is equivalent to minimizing $\mathcal{W}_\alpha$, which
has a simpler form. Thus, our clustering problem corresponds to
find the best partition of $\mathcal{X}$ such that
\begin{equation}\label{eq:minimize}
\min_{\{ \mathcal{A}_k \} } \mathcal{W}_\alpha(\{ 
\mathcal{A}_1, \dotsc, \mathcal{A}_{K}
\})
\end{equation}
where each datapoint belongs to one and only one partition (hard assignments).

Based on the equivalence between semimetrics of negative type
and kernel functions \eqref{eq:gen_kernel}, and choosing $x_0 = 0$ for
simplicity (one might choose any other point in $\mathbb{R}^D$),
in the case of \eqref{eq:g_def} we have the associated kernel
\begin{equation}\label{eq:kernel}
k_\alpha(x,y) = \tfrac{1}{2} \left( 
\| x \|^\alpha + \| y \|^\alpha - \| x-y \|^\alpha 
\right) .
\end{equation}
Now we can write
\begin{equation}\label{eq:W2}
\mathcal{W}_\alpha(\{ \mathcal{A}_1,\dotsc,\mathcal{A}_K \})
= - \sum_{k=1}^{K} \dfrac{N_k}{2} \, \mathbb{E} \, 
k_\alpha(\mathcal{A}_k, \mathcal{A}_k) = 
- \sum_{k=1}^{K} \dfrac{1}{2 N_k} \sum_{x, x' \in \mathcal{A}_k} 
k_\alpha(x, x').
\end{equation}

Now let $Z \in \{ 0,1 \}^{N\times K}$ be a binary matrix such
that 
\begin{equation}
Z_{nk} = \begin{cases}
1 & \mbox{if $x_n \in \mathcal{A}_k$ } \\
0 & \mbox{otherwise.}
\end{cases}
\end{equation}
Notice that $D = Z^T Z = \diag( N_1, N_2, \dotsc, N_K )$ contains the number
of elements in each partition. Introducing the kernel matrix
$K^{\alpha} \in \mathbb{R}^{N\times N}$ such that
\begin{equation}
K^{\alpha}_{ij} = k_\alpha(x_i, x_j),
\end{equation}
we can write \eqref{eq:W2} as
$- \tfrac{1}{2} \Tr D^{-1} Z^\top K^\alpha Z $. Thus our optimization problem
\eqref{eq:minimize} can be written as
\begin{equation}\label{eq:qcqp}
\begin{aligned}
& \max_{Z} \Tr\left\{ \big( Z D^{-1/2}\big)^\top K^{\alpha} 
\big( ZD^{-1/2} \big) 
\right\} \\
&\mbox{s.t. $Z_{ij} \in \{0,1\}$, $\sum_{k=1}^K Z_{nk} = 1$, 
$\sum_{n=1}^N Z_{nk} = N_k$, and $D = Z^\top Z$}.
\end{aligned}
\end{equation}
This is a quadratic problem with integer constraints, which is usually
NP-hard. This problem has the same formulation as Kernel $K$-means, but
with our non-parametric kernel function given by \eqref{eq:kernel}.
